---
title: "Overview of Mozilla Finances"
author: "Frank Hecker"
date: "8/15/2020"
output:
  html_document:
    fig_width: 9
    fig_height: 4.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this document I provide a graphical overview of the combined
finances of the nonprofit Mozilla Foundation and the Mozilla
Corporation, its wholly-owned subsidiary.

For those readers unfamiliar with the [R statistical software][r] and
the additional [Tidyverse software][tv] I use to manipulate and plot
data, I’ve included some additional explanation of various steps.  For
more information check out the the tutorial “[Getting started with the
Tidyverse][gs]”.

[r]: https://www.r-project.org/
[tv]: https://www.tidyverse.org/
[gs]: https://www.datacamp.com/community/tutorials/tidyverse-tutorial-r

## Setup and data preparation

### Libraries

I use the following R packages for the following purposes:

* tidyverse: do general data manipulation.

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
```

### Data sources

I use data from the independent auditors’ reports and consolidated
financial statements for the combined Mozilla organizations for 2005
(the year in which the Mozilla Corporation was founded) to 2018 (the
latest report available at the time of writing). I extracted data from
those statements into the following files:

* `mozilla-revenues.csv`. CSV file of data from the financial
  statements relating to revenues coming in to the combined
  organizations from third parties.
* `mozilla-expenses.csv`. CSV file of data from the financial
  statements relating to expenses incurred by the combined
  organizations.

The files do _not_ contain any data relating to transfers between the
organizations, for example, royalties paid to the Mozilla Foundation
by the Mozilla Corporation for use of the Mozilla, Firefox, and other
trademarks, or money paid by the Mozilla Foundation to the Mozilla
Corporation for office space, etc. That data can be found in the IRS
Form 990 documents filed by the Mozilla Foundation to the US Internal
Revenue Service. Analyzing that data is a potential future project for
me or someone else.

The file `mozilla-revenues.csv` contains multiple fields as shown
below. All values are in USD. Positive contributions to revenue
(gains) are stored as positive numbers, while negative contributions
to revenue (losses) are stored as negative numbers.

* `revenue`. Total revenue from all sources. Reported as “Total
  revenue and support” for 2005 to 2008 and 2018 and as “Total
  unrestricted revenue and support” for 2009 to 2017.
* `royalties`. Royalties paid to the combined organizations by third
  parties. (This is not to be confused with royalties paid to the
  Mozilla Foundation by the Mozilla Corporation, which are considered
  inter-organizational transfers and are not reported on the
  consolidated financial statements.) Reported as “Royalties”
  beginning in 2008. For 2005 to 2007 this category was instead
  reported as two different categories, “Royalties - Search” and
  “Royalties - Product Sales”.
* `royalties_search`. Royalties from search engine providers. Reported
  as “Royalties - search” for 2005 to 2007. Not reported separately
  beginning in 2008.
* `royalties_product`. Royalties received in connection with product
  sales. Reported as “Royalties - product sales” for 2005 to 2007. Not
  reported separately beginning in 2008.
* `subs_ads`. Revenue from subscriptions and advertising. Reported as
  “Subscription and advertising revenue“ beginning in 2017. Not
  reported separately for 2005 to 2016.
* `product`. Revenue from products (as distinguished from product
  royalties). Reported as “Product revenues” for 2005 to 2007. Not
  reported separately beginning in 2008.
* `services`. Revenue from services provided under contracts with
  third parties.  Reported as “Contracted services” for 2005 to
  2007. Not reported separately beginning in 2008.
* `interest`. Interest and dividends received from third
  parties. Reported as “Interest income” for 2005 and 2006 and as
  “Interest and dividend income” beginning in 2007.
* `investments`. Gains or losses resulting from investments made by
  either organization. Reported variously as “Unrealized losses from
  investments“ (2005), “Unrealized gain (loss) from investments”
  (2006), “Net realized and unrealized gain from investments” (2007,
  2013--2014), “Net realized and unrealized (loss) gain from
  investments” (2008, 2011), “Net realized and unrealized loss from
  investments” (2009--2010), “Net realized and unrealized gain (loss)
  from investments“ (2012, 2015--2017), and “Net realized and
  unrealized (loss) gain on investments, net” (2018).
* `contributions`. Contributions received from third parties, e.g., as
  grants or donations to the Mozilla Foundation. Reported as
  “Contributions” for 2005 and 2006 and for 2008 and subsequent years,
  and as “Contributions - Unrestricted” in 2007.
* `other`. Income not included in any of the other
  categories. Reported as “Other income” for 2005 and 2006 and as
  “Other” beginning in 2015. Not reported separately for 2007 to 2014.
* `forex`. Gains or losses from foreign currency exchange. Reported as
   “Foreign currency exchange (loss) gain” beginning in 2009, except
   for 2017 when it was reported as “Foreign currency exchange gain
   (loss)”. Not reported separately for 2005 to 2008.
* `assets`. Losses resulting from the sale of assets. Reported as
  “Loss on sales of assets” beginning in 2009. Not reported separately
  for 2005 to 2008.
* `released`. Assets released from previous restrictions and made
  available for use on an unrestricted basis. (For example, this
  apparently applies to certain grants made to the Mozilla Foundation
  by third parties.) Reported as “Net assets released from
  restrictions” beginning in 2009. Not reported separately in 2005,
  2006, and 2008. The 2007 statements have a category for restricted
  contributions that is related to this.

The file `mozilla-expenses.csv` contains fields as shown below. All
values are in USD. Expenses are stored as positive numbers, so must be
subtracted from revenues to determine profits (or more correctly, net
gains in assets, since the Mozilla Foundation does not have profits as
such).

* `expenses`. Total expenses for all categories. Reported as “Total
  expenses”.
* `software`. Expenses related to software development. Reported as
  `Software development`.
* `programs`. Expenses related to various projects and programs
  undertaken in support of the overall nonprofit goals of the combined
  organizations, including events, fellowships, sponsorships, and
  grants made to third parties. Reported as “Program services” for
  2006 to 2017 and as “Other program services” in 2018.
* `depreciation_program`. Depreciation expenses related to software
  development and other program services. Reported as “Depreciation”
  under “Program” for 2013 to 2017. Not reported separately for 2018.
* `marketing`. Expenses related to sales, branding, and
  marketing. Reported as `Sales and marketing` for 2005 to 2007 and as
  “Branding and marketing” beginning in 2008.
* `general`. Expenses related to running the organizations, including
  employee salaries and benefits, facilities, etc. Reported as
  “General and administrative”.
* `depreciation_support`. Depreciation expenses related to marketing
  and general and administrative activities in support of the combined
  organizations. Reported as “Depreciation” under “Support” for 2013
  to 2017. Not reported separately for 2018.
* `fundraising`. Expenses related to raising funds, e.g., to solicit
  individual donations or grants from third-parties. Reported as
  “Fundraising and development” beginning in 2016.
* `depreciation_fundraising`. Depreciation expenses related to
  fundraising. Reported as “Depreciation” under “Fundraising” for 2016
  and 2017. Not reported separately for 2018.
* `other`. Other expenses not allocated to any of the categories above.
  Reported as “Other expenses” for 2011 and 2012.

### Reading in the data

I begin by reading in the two CSV files into the data frames `revenues`
and `expenses`.

```{r read_csv}
revenues <- read_csv("mozilla-revenues.csv")
expenses <- read_csv("mozilla-expenses.csv")
```

I then clean the data to make it easier to manipulate:

* I need to treat royalties specially: prior to 2008 royalties from
  search engine providers were reported separately from other
  royalties, while in 2008 and subsequent years they were
  combined. For years in which the `royalties` data is missing I
  calculate `royalties` as the sum of the product and search
  royalties. I then remove the data for product and search royalties
  as no longer needed.
* To make further calculations easier, I replace any remaining missing
  values with zero in both the revenue and expense tables.

```{r royalties}
revenues <- revenues %>%
  mutate(royalties = ifelse(is.na(royalties),
                            royalties_product + royalties_search,
                            royalties)) %>%
  select(-royalties_product, -royalties_search) %>%
  replace(is.na(.), 0)

expenses <- expenses %>%
  replace(is.na(.), 0)
```

## Analysis

## Mozilla revenue vs. expenses

I first create an overview showing overall revenues versus expenses
for the years covered by the data.

I first create a new table `r_vs_e` by joining the `revenues` and
`expenses` tables on their common field `year`, and then retaining
only the `year`, `revenue`, and `expenses` fields.

```{r r_vs_e}
r_vs_e <- inner_join(revenues, expenses, by = "year") %>%
  select(year, revenue, expenses)
```

I then plot the `revenue` and `expenses` data with `year` on the x-axis.

```{r ggplot}
r_vs_e %>%
  mutate(revenue = revenue / 1000000, expenses = expenses / 1000000) %>%
  pivot_longer(revenue:expenses) %>%
  ggplot() +
  geom_line(
    mapping = aes(x = year, y = value, group = name, color = name),
    size = 0.8
  ) +
  ylab("") +
  xlab("Year") +
  scale_x_continuous(breaks = seq(2002, 2018, 2)) +
  scale_y_continuous(breaks = seq(0, 600, 100),
                     labels = scales::dollar_format(suffix = "M")) +
  scale_color_manual(values = c("#E69F00", "#56B4E9"),
                     breaks = c("revenue", "expenses"),
                     labels = c("Revenue", "Expenses")) +
  labs(title="Mozilla Revenue vs. Expenses",
       subtitle="Mozilla Foundation and Mozilla Corporation Combined",
       caption=paste0(
         "Data source:",
         "\n    Mozilla Foundation and Subsidiary Independent Auditors’ Report and Consolidated Financial Statements, 2005-2018")) +
  theme_minimal() +
  theme(axis.title.x=element_text(margin=margin(t=10))) +
  theme(axis.title.y=element_text(margin=margin(r=10))) +
  theme(plot.caption=element_text(margin=margin(t=15), hjust=0)) +
  theme(legend.title = element_blank())
```

### Revenue breakdown over the years

I next want to look at broad categories of revenue and how they have
grown or shrunk over the years. To do this I categorize the various
types of revenue as follows:

* I keep royalties as a separate category, since it has been by far
  the major source of revenue over the years, with almost all of it
  coming from Google and other search engine providers.
* I also keep subscription and ad revenue as a separate category,
  since it presumably will be where new sources of revenue such as the
  Mozilla VPN get reported.
* I combine the `contributions` and `released` variables into a single
  variable `grants_donations`, based on my understanding that the
  `released` values primarily reflect grants to the Mozilla Foundation
  that have associated conditions placed on them, with the money not
  being released for unrestricted use until the conditions are
  satisfied.
* I combine all other sources of revenue (or losses) into a single
  category `misc_revenue`.

```{r revenue_categories}
revenue_categories <- revenues %>%
  mutate(grants_donations = contributions + released,
         misc_revenue = product + services + interest + investments + other + forex + assets) %>%
  select(year, revenue, royalties, subs_ads, grants_donations, misc_revenue)
```

I then plot the various categories of revenue for the years for which
I have data.

```{r plot_revenue_categories}
revenue_category_palette <- c("#000000", "#E69F00", "#56B4E9",
                              "#009E73", "#CC79A7")

revenue_categories %>%
  mutate(across(revenue:misc_revenue, ~.x / 1000000)) %>%
  pivot_longer(revenue:misc_revenue) %>%
  ggplot() +
  geom_line(
    mapping = aes(x = year, y = value, group = name, color = name),
    size = 0.8
  ) +
  ylab("") +
  xlab("Year") +
  scale_x_continuous(breaks = seq(2002, 2018, 2)) +
  scale_y_continuous(breaks = seq(0, 600, 100),
                     labels = scales::dollar_format(suffix = "M")) +
  scale_color_manual(values = revenue_category_palette,
                     breaks = c("revenue", "royalties", "subs_ads",
                                "grants_donations", "misc_revenue"),
                     labels = c("Total Revenue",
                                "Royalties",
                                "Subscriptions and advertising",
                                "Grants and donations",
                                "Miscellaneous")) +
  labs(title="Mozilla Revenue Categories",
       subtitle="Mozilla Foundation and Mozilla Corporation Combined",
       caption=paste0(
         "Data source:",
         "\n    Mozilla Foundation and Subsidiary Independent Auditors’ Report and Consolidated Financial Statements, 2005-2018")) +
  theme_minimal() +
  theme(axis.title.x=element_text(margin=margin(t=10))) +
  theme(axis.title.y=element_text(margin=margin(r=10))) +
  theme(plot.caption=element_text(margin=margin(t=15), hjust=0)) +
  theme(legend.title = element_blank())
```

I then plot the revenue categories as a percentage of total revenue.

```{r plot_revenue_categories_pct}
revenue_categories %>%
  mutate(across(revenue:misc_revenue, ~100 * .x / revenue)) %>%
  pivot_longer(revenue:misc_revenue) %>%
  ggplot() +
  geom_line(
    mapping = aes(x = year, y = value, group = name, color = name),
    size = 0.8
  ) +
  ylab("") +
  xlab("Year") +
  scale_x_continuous(breaks = seq(2002, 2018, 2)) +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = scales::percent_format(scale = 1)) +
  scale_color_manual(values = revenue_category_palette,
                     breaks = c("revenue", "royalties", "subs_ads",
                                "grants_donations", "misc_revenue"),
                     labels = c("Total Revenue",
                                "Royalties",
                                "Subscriptions and advertising",
                                "Grants and donations",
                                "Miscellaneous")) +
  labs(title="Mozilla Revenue Categories as a Percentage of Total Revenue",
       subtitle="Mozilla Foundation and Mozilla Corporation Combined",
       caption=paste0(
         "Data source:",
         "\n    Mozilla Foundation and Subsidiary Independent Auditors’ Report and Consolidated Financial Statements, 2005-2018")) +
  theme_minimal() +
  theme(axis.title.x=element_text(margin=margin(t=10))) +
  theme(axis.title.y=element_text(margin=margin(r=10))) +
  theme(plot.caption=element_text(margin=margin(t=15), hjust=0)) +
  theme(legend.title = element_blank())
```

### Expense breakdown over the years

I next want to look at broad categories of expenses and how they have
grown or shrunk over the years. To do this I categorize the various
types of expenses as follows:

* I keep expenses related to software development (`software`) as a
  separate category, since it represents the main activity of the
  combined organizations and reflects value provided by Mozilla to the
  world at large in the form of open source software products.
* I also keep program-related expenses (`programs`) as a separate
  category, since it represents another form of value provided by
  Mozilla to the world at large, in the form of grants, fellowships,
  sponsorships, events, and so on.
* I retain marketing expenses, general and administrative expenses,
  and fundraising expensese as their own categories.
* I combine all other expenses, including depreciation-related
  expenses, into a single category `misc_expenses`. (I wasn’t sure
  exactly what to do with depreciation expenses, since as reported
  they can’t be exactly assigned to the various other categories.)

```{r expense_categories}
expense_categories <- expenses %>%
  mutate(misc_expenses = depreciation_program + depreciation_support +
           depreciation_fundraising + other) %>%
  select(year, expenses, programs, software, marketing, general,
         fundraising, misc_expenses)
```

I then plot the various categories of expenses for the years for which
I have data.

```{r plot_expense_categories}
expense_category_palette <- c("#000000", "#E69F00", "#56B4E9", "#009E73",
                              "#0072B2", "#D55E00", "#CC79A7")

expense_categories %>%
  mutate(across(expenses:misc_expenses, ~.x / 1000000)) %>%
  pivot_longer(expenses:misc_expenses) %>%
  ggplot() +
  geom_line(
    mapping = aes(x = year, y = value, group = name, color = name),
    size = 0.8
  ) +
  ylab("") +
  xlab("Year") +
  scale_x_continuous(breaks = seq(2002, 2018, 2)) +
  scale_y_continuous(breaks = seq(0, 600, 100),
                     labels = scales::dollar_format(suffix = "M")) +
  scale_color_manual(values = expense_category_palette,
                     breaks = c("expenses", "software", "general",
                                "marketing", "programs", "fundraising",
                                "misc_expenses"),
                     labels = c("Total Expenses",
                                "Software development",
                                "General and administrative",
                                "Marketing",
                                "Programs",
                                "Fundraising",
                                "Miscellaneous Expenses")) +
  labs(title="Mozilla Expense Categories",
       subtitle="Mozilla Foundation and Mozilla Corporation Combined",
       caption=paste0(
         "Data source:",
         "\n    Mozilla Foundation and Subsidiary Independent Auditors’ Report and Consolidated Financial Statements, 2005-2018")) +
  theme_minimal() +
  theme(axis.title.x=element_text(margin=margin(t=10))) +
  theme(axis.title.y=element_text(margin=margin(r=10))) +
  theme(plot.caption=element_text(margin=margin(t=15), hjust=0)) +
  theme(legend.title = element_blank())
```

I then plot the expense categories as a percentage of total expenses:

```{r plot_expense_categories_pct}
expense_categories %>%
  mutate(across(expenses:misc_expenses, ~100 * .x / expenses)) %>%
  pivot_longer(expenses:misc_expenses) %>%
  ggplot() +
  geom_line(
    mapping = aes(x = year, y = value, group = name, color = name),
    size = 0.8
  ) +
  ylab("") +
  xlab("Year") +
  scale_x_continuous(breaks = seq(2002, 2018, 2)) +
  scale_y_continuous(breaks = seq(0, 100, 20),
                     labels = scales::percent_format(scale = 1)) +
  scale_color_manual(values = expense_category_palette,
                     breaks = c("expenses", "software", "general",
                                "marketing", "programs", "fundraising",
                                "misc_expenses"),
                     labels = c("Total Expenses",
                                "Software development",
                                "General and administrative",
                                "Marketing",
                                "Programs",
                                "Fundraising",
                                "Miscellaneous Expenses")) +
  labs(title="Mozilla Expense Categories as a Percentage of Total Expenses",
       subtitle="Mozilla Foundation and Mozilla Corporation Combined",
       caption=paste0(
         "Data source:",
         "\n    Mozilla Foundation and Subsidiary Independent Auditors’ Report and Consolidated Financial Statements, 2005-2018")) +
  theme_minimal() +
  theme(axis.title.x=element_text(margin=margin(t=10))) +
  theme(axis.title.y=element_text(margin=margin(r=10))) +
  theme(plot.caption=element_text(margin=margin(t=15), hjust=0)) +
  theme(legend.title = element_blank())
```

### Environment

I used the following R environment in doing the analysis above:

```{r sessionInfo}
sessionInfo()
```

### Source code

You can find the source code and data for this analysis at my
[mozilla-finances public code repository][mf]. This document and its
associated source code and data are available for unrestricted use,
distribution and modification under the terms of the [Creative Commons
CC0 1.0 Universal (CC0 1.0) Public Domain Dedication][cc0].  Stated
more simply, you’re free to do whatever you’d like with it.

[mf]: https://github.com/frankhecker/mozilla-finances
[cc0]: https://creativecommons.org/publicdomain/zero/1.0/
